---
title: "Awesome Iris Data"
author: "Heath Blackmon"
date: 'last updated: `r Sys.Date()`'
output:
  html_document:
  theme: cosmo
---
  
load the iris dataset

```{r}
data(iris)
```

Check out the data and make sure we understand its structure.

```{r}
head(iris)
```

We want to calculate a value for every row so lets create a vector for those results.
```{r}
result <- vector()
```

Now we are ready to use a control structure and iterate through a dataset.

```{r}
for(i in 1:nrow(iris)){
  result[i] <- iris$Petal.Width[i]/iris$Petal.Length[i]
}
```



lets plot the results:
```{r}
plot(result)
```

That is interesting looks like there might be unique different groups.  If we look at the last column we see that there are actually 3 species.  The species are in grouped species 1 is on rows 1-50 two is 51-100 and the last one is on rows 101-150.  Lets draw some lines to separate the species and see if that makes sense.

```{r}
plot(result)
abline(v=50, col="red")
abline(v=100, col="red")
```
