---
title: "Things to explore later"
author: "Heath Blackmon"
output: 
  html_document:
    theme: cosmo
    highlight: kate
    toc: true
    toc_depth: 2
---


[Return to main website](http://coleoguy.github.io/resources.html)

___
#Distribution of subtree sizes.

The distribution of sub-tree sizes seems to be largely a function of tree balance.  I believe a lot of work was done on this in the 80s but not sure.  For instance  a completely ladderized tree and a completely symmetric tree have very different distributions.

```{r}
par(mfcol=(c(1,2)))
phy <- ape::read.tree(text = "(H:7,(G:6,(F:5,(E:4,(D:3,(C:2,(A:1,B:1):1):1):1):1):1):1);")
phy2 <- ape::read.tree(text = "(((A:1,B:1):1,(C:1,D:1):1):1,((E:1,F:1):1,(G:1,H:1):1):1);")
plot(phy)
plot(phy2)
phycount <- 2:128
phy2count <- c(128,rep(64,2),rep(32,4),rep(16,8),rep(8,16),rep(4,32),rep(2,64))
```

If we think about trees like this but expand it out to say 128 tips we can start to get an idea of the distributions of sub-trees:

```{r}
plot(density(phycount), main="ladderized tree", xlab="subtree sizes")
plot(density(phy2count), main="symmetric tree", xlab="subtree sizes")
```

So does this difference tell us anything about the generating process?  Tonjana has probably looked at this a little bit should probably talk to him about it.


