---
title: "Bayes Theorem Example"
author: "Heath Blackmon"
date: "1/30/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Example Problem
Imagine that we have a disease that is preseint in 1 in a million people in the general population. We have developed a test for this disease that is 99.9% correct in returning a positive result when someone is truly infected. However, it has a false positive rate of 5%. Lets use Bayes Theorem to calculate the probility that a random individual chosen from the population that tests positive actually  has the disease.

## Bayes Theorem
We will use $hyp$ to represent our hypothesis that someone has the disease and we will use $post$ to represent a positive test. Also remember that $p$ is read as probability and the $|$ character is read as given that. So the equation $P(hyp|post)= \frac{P(post|hyp)P(hyp)}{P(post)}$ would be read as the probability that a person has the disease given that they tested positive is equal to the probability that you get a positive test given that someone has the disease times the probability of someone having the disease (this is the prior) divided by the probability of getting a positive test. Remember that there are two kinds of positive tests: true positives and false positives and both must be accounted for.

Lets use our quickly building R skills to work out this example question in R. I will use the variables posthyp, hyp, and post to hold the probabilities of each of the 3 values on the right side of Bayes Theorem

```{r}
# first the probability of getting positive test given the test subject is infected
posthyp <- .999

# then the probability of a random person being infected
hyp <- 1/1000000

# the the probability of a positive test
false.pos <- .05 * 999999/1000000
true.pos <- .999 * 1/1000000
post <- false.pos + true.pos

# now we combine these in Bayes Theorem
prob.having.disease.if.test.positive <- posthyp * hyp / post

```


The chance that a random person who tests positive has the disease is `r prob.having.disease.if.test.positive`.

