---
title: "Open Science"
author: "Heath Blackmon"
date: 'last compiled: `r Sys.Date()`'
output:
  html_document:
    theme: cosmo
    toc: yes
---
# Loading data
R comes with many cool datasets; lets use the `chickwts` dataset today. In these experiment chicks were randomly allocated into groups, and each group was given a different feed supplement.
```{r}
data("chickwts")
```

# Data exploration and cleaning
First lets look at what this dataset looks like:
```{r}
knitr::kable(head(chickwts))
```

OK now lets look and see what the distribution of all chick weights looks like:
```{r}
hist(chickwts$weight)
```

That looks like something interesting might be going on. Lets try plotting this parsed by the type of feed. The `boxplot` has this built in using the `~` symbol to specify what we are parsing by. 
```{r}
boxplot(weight ~ feed, data = chickwts, col = "lightgray",
main = "chickwt data", ylab = "Weight at six weeks (gm)")
```

# Figures for manuscript
Now that we know what our data looks like lets produce figures for a manuscript. First we will use the `pdf` command this creates and opens the pdf file and that subsequent plotting commands are sent to. Once are plot is complete then we use the `dev.off` command to close the file:

this will be our figure 1 that is the distribution of all chick weights at 6 weeks:
```{r}
pdf(file="fig1.pdf", width = 4, height = 4)
hist(chickwts$weight, main = "Chick weights at 6 weeks",
xlab = "grams", ylab = "count")
dev.off()
```

next we will make our figure 2 that is the chick weights at 6 weeks parsed by feed type:
```{r}
pdf(file="fig2.pdf", width = 4, height = 4)
boxplot(weight ~ feed, data = chickwts, col = "lightgray",
main = "chickwt data", cex.axis=.5, ylab = "Weight at six weeks (gm)")
dev.off()
```