---
title: "Bayes Theorem"
author: "Heath Blackmon"
date: "3/8/2019"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bayes Theorem and Probability

Baye's theorem is an amazing tool that we can use to understand how much based on all of our data and previous knowledge how much we should believe in a hyothesis. The basic form of the theorem is:

$$p(A|B) = \frac{p(B|A) \times p(A)}{p(B)}$$

A represents our hypothesis while B represents our data. Lets apply this to a biological example.


Suppose you are studying a pair of cryptic species. In your area years of study show that 5% of individuals are species A and 95% of individuals are species B. There is currently no genetic assay capable of telling them apart. They differ however in the frequency of a rare color pattern. Species A has the rare color pattern 50% of the time while species B has the rare color pattern only 2% of the time.  Assume these numbers are known with certainty, from many years of field research. 

```{r}
# probabilities in the field
p_a <- .05
p_b <- .95

# conditional probabilities for color give species
p_rc_a <- .5
p_rc_b <- .02
```

### Problem 1
Now suppose you find one of these species with the rare color pattern. Use Bayes theorem to compute the probability that it is from species A.

So here is our basic equation for this problem:
$$p(species A | rare color) = \frac{p(rare color | species A) \times p(species A)}{p(rare color)}$$

```{r}
p_a_rc <- (p_rc_a * p_a) / (p_rc_a * p_a + p_rc_b * p_b)
p_a_rc
```

So we have a 56.8% chance that this specimen belongs to species A.

### Problem 2

A new paper comes out that states that species B has a morphological feature 50% time that is only present 10% of the time in species A.  Your sample has this feature.  What is the probability that you have a sample of species A now? 

### Problem 3

You believe that Thylacinus cynocephalus (a marsupial wolf analog), though widely believed to have gone extinct in 1933, may still be alive in thinly settled regions of Tasmania.  Using museum specimens, you have developed a DNA genotyping test that will accurately identify T. cynocephalus DNA 99.9% of the time.  You also evaluate your test by running 1000 marsupial samples that are not from the genus Thylacinus.  You find that 2 of these returned false positives. You go to Tasmania and collect scat, hair samples and bone fragments (all fresh) from a number of locations to test for Thylacinus DNA.  After testing all 73 of your samples you find that one of them are positive for Thylacinus DNA.  

Use Bayes theorem to determine the probability that you have identified the remains of a Thylacinus.

Justify your choice of a prior in this problem. 
