---
title: "Implementing Pagel's test in diversitree:"
author: "Heath Blackmon"
date: "September 19, 2015"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
---

[Return to main website](http://coleoguy.github.io/resources.html)

Return to of comparative methods 
[outline](http://coleoguy.github.io/comparative/methods.html) page 

```{r, echo=F}
load("~/Desktop/pagel.data.sets.RData")
num <- 3
phy <- set.phy[[num]]
dd1 <- dependant.data1[[num]]
dd2 <- dependant.data2[[num]]
id1 <- independant.data1[[num]]
id2 <- independant.data2[[num]]
rb1 <- perfrep1[[num]]
rb2 <- perfrep2[[num]]
ds1 <- state.darwin1[[num]]
ds2 <- state.darwin2[[num]]
ub1 <- unrepburst1[[num]]
ub2 <- unrepburst2[[num]]
rm(dependant.data1, dependant.data2, independant.data1, independant.data2,
   perfrep1, perfrep2, set.phy, state.darwin1, state.darwin2, transitioned1,
   transitioned2, unrepburst1, unrepburst2)
```

___
# Starting points in diversitree

One issue with using diversitree is the need to pick a starting point for the parameters prior to optimization.  We could just try multiple times picking numbers from a random uniform distribution that ranges from zero to some very large number. In fact [Dan Rabosky](http://www-personal.umich.edu/~drabosky/Home.html) even has a function in the supplemental materials from his paper with [Emma Goldberg](http://eeg.github.io/lab/home.html) [*Model Inadequacy and Mistaken Inferences of Trait-Dependent Speciation*](http://sysbio.oxfordjournals.org/content/64/2/340.short) that does just this.

It feels inelligant to pick some number that you "think" is to big.  We could do this in a bit more of a quantitative method by picking a max value based on the kinds of branch length present in out tree.
```{r}
# first lets find the average length of a terminal branch
avg.branch <-mean(phy$edge.length[which(!phy$edge[,2] %in% phy$edge[,1])])
```

Now lets look at the way that information about the base of a branch with a length of `avg.branch ` decays as the rate of evolution increases.

```{r}
x <- y <- vector()
for(i in 1:1000){
  rate <- i/100
  Q <- matrix(rate,2,2)
  diag(Q) <- -rate
  P <- ape::matexpo(Q*avg.branch)
  y[i] <- P[2,1]
  x[i] <- i/100
}
plot(x,y, pch=16,cex=.3,xlab="Rate in Q-matrix", ylab="Prob. of root state 0", main=paste("avg. terminal branch=", round(avg.branch,3), sep=""))

```

I think that looking at a plot like this for your dataset would be a good idea. It would let you pick a good value for the high end of possible starting points.

___
#Implementation of Pagel's Test [1]
```{r}
StartMK <- function(np){
		x <- runif(np, 0, 5);
	return(x);
}
lrt <- function (dep, ind) {
  L.dep <- dep$value
  L.ind <- ind$value
  L01 <- as.vector(- 2 * (L.dep - L.ind))
  df <- length(ind$par) - length(dep$par)
  pvalue <- pchisq(L01, df, lower.tail = FALSE)
  return(pvalue)
}
	



Pagel94 <- function(phy, x, y, maxbad=5000, opt=5){
  # first we have to combine our two binary traits into one 4 state trait
  z <- paste(x, y, sep="")
  z[z=="00"] <- 1
  z[z=="01"] <- 2
  z[z=="10"] <- 3
  z[z=="11"] <- 4
  z <- as.numeric(z)
  names(z) <- phy$tip.label
  
  lk1 <- diversitree::make.mkn(phy, z, k=4, strict=F)
  lk.ind <- diversitree::constrain(lk1, q14~0, q23~0, q32~0, q41~0) #pars=8
  lk.dep <- diversitree::constrain(lk.ind, q13~q24, q31~q42, q21~q43, q12~q34) #pars=4
  #  lk.ind(StartMK(8))
  #  lk.dep(StartMK(4))
	minb <- rep(0, 8)
	maxb <- rep(50, 8)
	
	# Fit the independent model
	for (i in 1:opt){
		badcount <- 0
		pp <- StartMK(8)
		resx <- try(optim(pp, lk.ind, method='L-BFGS-B', 
		                  lower=minb, upper=maxb, 
		                  control=list(maxit=5000, fnscale=-1)), silent=T);
		while (class(resx) == 'try-error'){
		  pp <- StartMK(8)
		  resx <- try(optim(pp, lk.ind, method='L-BFGS-B', 
		                    lower=minb, upper=maxb, 
		                    control=list(maxit=5000, fnscale=-1)), silent=T);
			if (badcount > maxbad){
				stop("too many fails to optimize independant model")
			}
			badcount <- badcount + 1
		}
 
		if (i == 1){
			best.ind <- resx
		}else{
			if (best.ind$value < resx$value){
				best.ind <- resx
			}
		}
	}

	minb <- rep(0, 4)
	maxb <- rep(50, 4)
	
	# Fit the dependent model
	
	for (i in 1:opt){
		badcount <- 0
		pp <- StartMK(4)
		resx <- try(optim(pp, lk.dep, method='L-BFGS-B', 
		                  lower=minb, upper=maxb, 
		                  control=list(maxit=5000, fnscale=-1)), silent=T);
		while (class(resx) == 'try-error'){
		  pp <- StartMK(4);
		  resx <- try(optim(pp, lk.dep, method='L-BFGS-B', 
		                    lower=minb, upper=maxb, 
		                    control=list(maxit=5000, fnscale=-1)), silent=T);
			if (badcount > maxbad){
				stop("too many fails to optimize dependant model")
			}
			badcount <- badcount + 1;
		}
		if (i == 1){
			best.dep <- resx;
		}else{
			if (best.dep$value < resx$value){
				best.dep <- resx;
			}
		}
	}
  m.comp <- lrt(best.dep, best.ind)
  return(m.comp)
}
```
lrt(foo.dep,foo.ind)
```{r}
mark(phy, x=dd1, y=dd2)
mark(phy, x=ds1, y=ds2)
mark(phy, x=id1, y=id2)
mark(phy, x=rb1, y=rb2)
mark(phy, x=ub1, y=ub2)
Pagel94(phy, x=dd1, y=dd2)
Pagel94(phy, x=ds1, y=ds2)
Pagel94(phy, x=id1, y=id2)
Pagel94(phy, x=rb1, y=rb2)
mark(phy, x=ub1, y=ub2)
```

___
#citations
[1] Pagel, Mark. "Detecting correlated evolution on phylogenies: a general method for the comparative analysis of discrete characters." Proceedings of the Royal Society of London B: Biological Sciences 255, no. 1342 (1994): 37-45.

[2] Huelsenbeck, John P., Rasmus Nielsen, and Jonathan P. Bollback. "Stochastic mapping of morphological characters." Systematic Biology 52, no. 2 (2003): 131-158.

[3] Maddison, Wayne P. "A method for testing the correlated evolution of two binary characters: are gains or losses concentrated on certain branches of a phylogenetic tree?." Evolution (1990): 539-557.


