---
title: "Bayes Homework - Solved"
author: "Heath Blackmon"
date: "2/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem
Suppose you are studying a pair of cryptic species. In your area 5% of individuals are species A and 95% of individuals are species B. There is currently no genetic assay capable of telling them apart. They differ however in the frequency of a rare color pattern. Species A has the rare color pattern 50% of the time while species B has the rare color pattern only 2% of the time. Assume these numbers are known with certainty, from many years of field research. 

## Question 1
Now suppose you find one of these species with the rare color pattern. Use Bayes theorem to compute the probability that it is from species A.

## Bayes Theorem Refresher
We will use $hyp$ to represent our hypothesis that our specimen is species A and we will use $post$ to represent having the color pattern that is more common in rare species. So the equation $P(hyp|post)= \frac{P(post|hyp)P(hyp)}{P(post)}$ would be read as the probability that a specimen is species A given that it has the color pattern common in this species is equal to the probability that this color pattern is present given that it is species A times the probability of species A divided by the probability of this color pattern. 


```{r}
posthyp <- .5
hyp <- 5/100
false.pos <- .05 * .5
true.pos <- .95 * .02
post <- false.pos + true.pos

# now we combine these in Bayes Theorem
answer <- posthyp * hyp / post

```


The chance that our specimen is species A is `r round(answer, 3)`.


## Question 2

A new paper comes out that states that species B has a morphological feature 50% time that is only present 10% of the time in species A. Your sample has this feature. What is the probability that you have a sample of species A now?

Some of you took this as an idependent question and some of you combined infromation from both 1 and 2 together. That approach of combining all data is particularly appropriate in this case because the beauty of Bayes theorem is our ability to combine information from multiple resources into out prior. I'll accept either approach for the purposes of the homework.

```{r}
# with only this info

# with all avaialable info
```



## Question 3

Suppose now that a genetic test is developed that can identify the species of our sample. But the test, like all tests, is imperfect. This is the information you have about the test:
a. The probability it correctly identifies species A is 0.7
b. The probability it correctly identifies species B is 0.98

You run the test and it is positive for species A. Compute the posterior probability that your sample is
species A using all the information available. Again I'll show both approaches with just this info and with all information combined.


```{r}
# with only this info

# with all avaialable info
```





