---
title: "Sampling and Summary Statistics"
subtitle: "Biology 683 - Lecture 2"
author: "Heath Blackmon"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  ioslides_presentation:
    incremental: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Last Week

- What are some causes of the reproducibility crisis?  
- What makes R awesome?  






## Today

- Terminology  
- Summarizing Data  
- Central Limit Theorem  
- Covariance and Correlation  






## Populations and Samples

- **Populations**  
- Some sort of group of something - could be anything  

		- Undergraduates at Texas A&M   
		
		- Jewel beetles in Arizona   
		
		- Strain of flies in the lab   

- **Samples**  
- A subset of individuals drawn from a population 


## What is the population?

*We wanted to examine any association between the severity of injuries, and the height from which cats fall in high-rise buildings.*  

*In the period between January 1, 1998 and December 12, 2001 at the Clinic of Surgery, Orthopaedics and Ophthalmology of the Veterinary Faculty, 119 cats were treated after a fall or jump from a balcony or window, where the owners saw the fall, or where there was a reasonable suspicion that a fall had occurred. Only those cats that fell from the second or higher stories were included. The owners brought the cats for treatment within varying periods of time after the fall (from 30 min to over a month).*


<font size="4"> Vnuk, et al. "Feline high-rise syndrome: 119 cases (1998–2001). *Journal of Feline Medicine & Surgery* 6.5 (2004): 305-312.</font> 

## Sampling Considerations

- **Target population**
- Need to sample a representative population  
- A sample of people from College Station, for instance, would probably not be representative of New Yorkers 

- **Sampling Error** 
- Chance alone will cause your sample to depart from the population



## Parameter, estimates, sampling considerations

- **Parameter**  
- Population-level variables we are trying to estimate   

- **Estimate or Statistic**   
- The value of the parameter inferred from the sample   

- **Bias**  
- If something about the sampling procedure causes the sample to systematically misrepresent the population 

- **Precision**  
- How tightly grouped are the estimates? 





## Accuracy vs Precision

```{r echo=FALSE}
library(png)
library(grid)
img <- readPNG("images/acc.prec.png")
grid.raster(img)
```

- Precision is a measure of spread  
- Accuracy is a measure of bias  





## Random Sampling

- Every unit in a population should have an equal chance of being sampled  
- The selection of units must be independent   







## Your big idea should be a hypothesis

- A statistical hypothesis is a specific claim about a 	population parameter 
- *Caloric restriction increases the lifespan of Drosophila melanogaster*  
- *Pesticide exposure causes feminization of amphibian males*






## Data

- **Variables**  
- The characteristics that differ among individuals 

- **Data**   
- The measurements of variables taken for a sample of individuals 

- **Categorical Variables**  
- Individuals are in qualitative categories 





## Data

- **Numerical Variables** 
- Individuals vary on a quantitative scale
 
- **Ordinal** 
- The categories can be ordered 

- **Nominal** 
- The categories have no inherent order








## Continuous vs Discrete

- **Continuous variables**  
- a variable that has an infinite number of possible values  

- **Discrete variables**  
- a variable that has a finite number of possible values  





## Explanatory and Response Variables




## For Thursday

Read chapters 1 and 2 of WS

Install R and Rstudio on a laptop  
-	Come and see me before class on   
-	Thursday if you run into problems  

Bring laptop to class!

Heath Blackmon  
BSBW 309  
coleoguy@gmail.com  
@coleoguy  





